---
title: "Introduction to `data.table`"
author: "Ted Laderas"
date: "July 20, 2016"
output: html_document
---

#1.0 A Quick Introduction to the `data.table` package

We are using the `data.tables` package to store the data in memory. There are many advantages of a `data.table` versus using the regular `data.frame`:

1. Subsetting and doing aggregate calculations (such as you would use `tapply()` for) are very fast for large datasets. This is because `data.table` avoids many of the memory-copying problems of current base R functions.
2. The `fread()` function to load in data in delimited files is very fast, much faster than the base function `read.table()`.
3. Joining tables is very fast and efficient.

The main disadvantage of `data.table` is that the syntax is different than for `data.frames` or `dplyr`. It is just different enough to drive you crazy. The basic syntax is as follows:

```{r eval=FALSE}
DT[i,j,by]
```

Where 
- `i` is an operation/function on rows, to produce row filtering (also called an `i`-statement),
- `j` is an operation/function on columns, to calculate values on columnn (also called a `j`-statement),
- `by` is an optional argument that specifies a column to aggregate on.

We'll slowly go over all of these operations.

##1.1 Keys For `data.table`

A `data.table` has a key, which you can set by using the `setkey()` function. This key is usually a column name, though you can also use multiple columns here. This key has two purposes: it provides an *index* to sort the `data.table`, and for joining with other tables, it provides the identifier to join on. Joins/merges with data.table are very quick.

```{r}
#make a data.table version of iris data
data(iris)

irisDT <- data.table(iris)

setkey(irisDT, Sepal.Length)
irisDT
```

##6.2 Merges/Joins of Two `data.tables`

Let's specify another table to join `irisDT` on. To do so, we'll initialize a brand new `data.table` called `testTable`.

```{r}
#make a little table to merge
#initializing a data.table is identical to initializing a data.frame
testTable <- data.table(Species=c("setosa", "versicolor", "virginica"), 
                        Likes=c(TRUE,FALSE,FALSE), Color = c("purple", "purple", "pink"))
#show the table
testTable
```

Now we need to set common keys for both tables to join them.

```{r}
setkey(irisDT, Species)
irisDT

setkey(testTable, Species)
```

Now that we have the keys set for both tables, we can join them together. In `data.table`, this is called a *rolling join*, and it is very fast, even for datasets with millions of rows. 

```{r}
#merge the two tables together using the keys we set (Species)
irisDT[testTable]
```

You may notice that the syntax here is completely different than `data.frame`. The way to read the above merge is that we want to merge the rows of `irisDT` and `testTable` given the keys that we have specified for each table.

###1.2 Casting the long data into wide data

If we wanted the data in a `wide` matrix format with each row belonging to a probeset/patient/study triplet, and each column corresponding to the value of the time series, we can transform the data using `dcast`. This function lets us *reshape* the data into another format. 

The key to understanding `dcast` is understanding the formula interface for R.

```{r}
load("data/twoStudies.RData")

dcast(viralData, FEATUREID + SUBJECTID + STUDYID ~ TIMEHOURS, value.var= "value", fun.aggregate = mean)
```

##1.3 Subsetting Rows

This operation is very fast, and is the main reason we're using `data.tables`. Let's look at the first entry in the `pathways` object.

```{r}
#get the name of the first list item
pway <- names(pathways)[1]
pway

path1 <- pathways[[pway]]
path1
```

##1.4 Subsetting Columns

You can subset columns the `data.frame` way by using the optional `with=FALSE` argument. This is important to know if you're passing a character vector to subset the data.

```{r}
irisDT[,c("Species", "Sepal.Length"), with=FALSE]
```

The `data.table` way is to pass a list of values wrapped up in `.()`, which is a shortcut for list. Also note that we don't quote the names of the columns we ask for.

```{r}
irisDT[,.(Species, Sepal.Length)]
```

##1.5 Running Operations on Columns

Within the `j`-statement, we can run all sorts of functions. For example:

```{r}
irisDT[,mean(Sepal.Width)]
```

Much like the `tapply()` command can be run on `data.frames` to aggregate data, we can specify operations or functions on columns (which are called *`j`-statements*) and use the `by=` argument to specify the operation. Here, we say that we want to aggregate the values by `Species` and produce  

```{r}
irisDT[,mean(Sepal.Width), by=Species]
```

If we want to return multiple outputs, we have to wrap our `j`-statement with the `.()` alias. Within these alias, we can also name the output columns.

```{r}
irisDT[,.(meanSW=mean(Sepal.Width), sdSW=sd(Sepal.Width)), by=Species]
```

##1.6 Calculating New Values on Columns

One of the most powerful operations is of `data.table` is that you can operate on columns in place using the `:=` operator. 

Obviously, this can be dangerous if you are rewriting columns, so take caution with this.

##1.7 Putting it all together
