---
title: "Shiny/EDA Tutorial"
author: "Ted Laderas"
date: "July 18, 2016"
output: html_document
---

The goal of today's workshop is to examine a set of genes across a number of cohorts for Dream Challenge Dataset.

There are seven studies in total with the following characteristics. The most important is the Duke Rhinovirus dataset, which contained patients who were exposed to Rhinovirus 

##Before You Start

Please make sure that you have the following packages installed:

```{r eval=FALSE}
library(data.tables)
library(ggplot2)
library(dplyr)
library(shiny)
```

##Filtering 

The set of 20,000 probesets from the Affymetrix U133 2.0 microarray were mapped to the gene sets and filtered using the following approach. The Coefficients of Variation for the Sham (non exposed) patients at all timepoints were compared to the CVs for those patients exposed to Rhinovirus. The thought behind this is that genes of interest should show more variability (due to the time-series design) in the Rhinovirus patients than the sham patients.

Based on visualizing the distributions of CVs, a cutoff of `cvRhinovirus/cvSham` > 1.4 was used, leading to a set of 836 genes that had higher variability in the Rhinovirus compared to Shams. We'll be visualizing this much smaller set across multiple patients, diseases, and time points.

```{r}
summary(cars)
```

##Using `data.tables`




##Visualizing Gene Sets



##Clustering The Data